---
title: "HW2 - DATA 609"
author: "Thomas Hill"
date: "September 18, 2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```





__Ex. 1__ -- Show that $x^2 + exp(x) + 2x^4 + 1$ is convex.  


__This function f(x) is convex if the following is true:__

$$f(\alpha x + \beta y) \leq \alpha f(x)  + \beta f(y), \alpha \geq 0, \beta \geq 0, \alpha + \beta = 1$$

__However, the consequent equation is time-consuming to simplify. One property of convex functions makes determining this easier:__

__If $f_1$ and $f_2$ are convex, then $\alpha f_1 + \beta f_2$ is also convex for any $\alpha, \beta \geq 0$.__  


__Consider $f_1 = x^2, $f_2 = 2x^4 + 1$, and $f_3 = exp(x)$:__

$$f_1: (\alpha x + \beta y)^2 \leq  \alpha x^2   +  \beta y^2$$
__A similar quadratic funciton has larady been solved in the lesson, but to recap:__

$$ 0 \leq  \alpha x^2   +  \beta y^2 - (\alpha x + \beta y)^2 $$

$$ 0 \leq \alpha x^2 - \alpha^2 x^2 + \beta y^2 - \beta ^2 y^2 - 2\alpha \beta xy$$
__Factor out $\alpha$ and $\beta$ terms:__
$$ 0 \leq \alpha x^2 ( 1 - \alpha ) + \beta y^2 (1 - \beta) - 2\alpha \beta xy = \alpha \beta x^2 + \alpha \beta y^2 - 2\alpha \beta xy $$

$$ \alpha \beta (x^2 - 2xy - y^2) = \alpha \beta (x - y)^2 \geq 0$$

__This inequality is always true for real numbers. Next, consider whether $exp(x)$ is convex:__  


$$f_2: exp(\alpha x + \beta y) \leq  \alpha exp(x)   +  \beta exp(y)$$

$$ 0 \leq  \alpha exp(x)   +  \beta exp(y) -  exp(\alpha x + \beta y)$$

__Divide by $exp(x)$, which is permissible as this can never be equal to zero:__  


$$ 0 \leq  \alpha   +  \beta exp(y - x) -  exp(\alpha x + \beta y - x) =  \alpha   +  \beta exp(y - x) -  exp(\beta y - x(1-\alpha))$$

$$ 0 \leq  =  \alpha   +  \beta exp(y - x) -  exp(\beta (y - x)) = \alpha -exp(\frac{(y-x)^\beta}{\beta(y-x)}) = \alpha + \alpha exp(\frac{(y-x)}{\beta})$$

__After much simplifying, we're left with two terms, both of which are non-negative for all real numbers.  Therefore exp(x) is convex.__


__Finally, $f_3 = x^4 + 1$ is convex owing to the property of convex functions as follows:__

__If f(x) and g(x) are convex, then f(g(x))s convex under non-decreasing conditions.  The f(x) and g(x) in this question are $f(x) = x^2$ and $g(x) = x^2 +1$.  __

__Taken all together, the original function is convex.__  



__Ex. 2__ -- Show that the mean of the exponential distribution

$$p(x) = \begin{cases} 
      \ \lambda e^{-\lambda x}, & 0 < \lambda \leq x \\
      0, & x <  0 
   \end{cases}$$

is $\mu = \frac{1}{\lambda}$ and its variance is $\sigma ^2 = \frac{1}{\lambda^2}$.  


__The mean $mu$ of a continuous probability distribution p(x) is defined as:__

$$ \mu = E(x) = \int x p(x) dx$$

__For the exponential distribution, this can be integrated by parts, e.g., $\int u dv = uv - \int v du$__


$$ \int ^\infty _0 x  \lambda e^{-\lambda x} = -x e^{-\lambda x} - \int e^{-\lambda x} = -e^{-\lambda x}(\frac{1}{\lambda}+x)|^\infty _0$$


__Using the limit of exp(-x) = 0 as x approaches infinity:__  


$$-\frac{e^{-\lambda x}}{\lambda}(1+x)|^\infty _0 =  0 + (\frac{1}{\lambda}+0) =  \frac{1}{\lambda} $$

__The variance $\sigma ^2| of a continuous probability distribution p(x) is defined as:__  


$$ Var(x) = E[x^2] = (\int x^2 f(x) dx) - \mu^2$$ 

__where $\mu$ is the mean. Substituting in for the exponential distribution:__  


$$ \int ^\infty _0 x^2 \lambda e^{-\lambda x} dx$$

__Using integration by parts again:__  



$$ \lambda \int ^\infty _0 x^2 e^{-\lambda x} dx  = -x^2 e^{-\lambda x} + 2\int x e^{-\lambda x} dx = -x^2 e^{-\lambda x} - \frac{2}{\lambda}e^{-\lambda x}(\frac{1}{\lambda}+x)|^\infty _0$$

$$-x^2 e^{-\lambda x}+ 2\frac{e^{-\lambda x}}{\lambda}(1+x)|^\infty _0 = 0 + 0 + 0 +\frac{2}{\lambda} *\frac{1} {\lambda} = \frac{2}{\lambda^2} $$

$$ \frac{2}{\lambda^2} - [\frac{1}{\lambda}]^2 = \frac{1}{\lambda ^2}$$

__The variance is equal to $\frac{1}{\lambda}$.__  


__Ex. 3__ -- It is estimated that there is a typo in every 250 data entries in a database, assuming the number of typos can obey the Poisson distribution. For a given 1000 data entries, what is the probability that there are exactly 4 typos? What is the probability of no typo at all? Use _R_ to draw 1000 samples with $\lambda = 4$ and show their histogram.  



__The Poisson distribution is $p(x) \frac{\lambda^x e^{-x}}{x!}$, with x is the number of events we're interested in observing, and $\lambda = np$, where n is the total number of occurences and p is the probability of an event. In this example, p = $\frac{1}{250}$, n = 1000 samples, and x = 4 typos.  The exact solution is:__  



```{r, poisson-calc1}

n = 1000
p = 1/250
x = 4

ex_3a <- ((n * p)^x * exp(-n*p))/factorial(x)

print(ex_3a)
```


__Approximately 0.195 of exactly 4 typos.__ 


__The probability of no typos is:__  



```{r, poisson-calc2}

n = 1000
p = 1/250
x = 0

ex_3b <- ((n * p)^x * exp(-n*p))/factorial(x)

print(ex_3b)
```

__Approximately 0.018.__

__Using a Poisson random variable, below is a histogram of 1000 samples of $\lambda = 1000 * \frac{1}{250} = 4$__  


```{r, poisson-sim}

set.seed(123)

hist(rpois(n = 1000, lambda = 1000/250))


```

__The most frequent outcome is approximately four typos.__  











