---
title: "HW6 - DATA 609"
author: "Thomas Hill"
date: "November 14, 2021"
output:
  html_document: default
  pdf_document: default
header-includes: -\usepackage{caption}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, library}

library(knitr)
library(class)


```



 

__Ex. 1__ -- Use a dataset such as the _PlantGrowth_ in R to calculate three distance metrics and discuss the results.



```{r, load-plant-growth}

pg <- PlantGrowth

plot(pg)
```

PlantGrowth is a dataframe comparing the yields of plants that had one of two treatments, or no treatment at all. Because these treatments only consider weight and another another dimension like treatment dose, I'll start by looking at the distance between points ranked by size. Because of this, the Euclidean and Manhattan distances are the same.

```{r, 1-d-norm}

sum(abs(sort(pg[pg$group == 'trt1',][['weight']]) - sort(pg[pg$group == 'ctrl',][['weight']])))


sum(abs(sort(pg[pg$group == 'trt2',][['weight']]) - sort(pg[pg$group == 'ctrl',][['weight']])))

```

Using this metric, treatment 1 appears slightly closer to the control.

Next, lets consider the distance of each treatment to its nearest control neighbor. I did this by subtracting each control value from a single treatment, found the absolute value of each difference, and located the minimum value of the vector. I did this for all 10 samples and summed the total distances as calculated by Euclidean distance.


```{r, 1-n-n}

pg_ctrl <- pg[pg$group == 'ctrl',][['weight']]
pg_t1 <- pg[pg$group == 'trt1',][['weight']]
pg_t2 <- pg[pg$group == 'trt2',][['weight']]


sum(min(abs(pg_t1[1] - pg_ctrl)) + min(abs(pg_t1[2] - pg_ctrl)) + min(abs(pg_t1[3] - pg_ctrl)) + min(abs(pg_t1[4] - pg_ctrl)) + min(abs(pg_t1[5] - pg_ctrl)) + min(abs(pg_t1[6] - pg_ctrl)) + min(abs(pg_t1[7] - pg_ctrl)) + min(abs(pg_t1[8] - pg_ctrl)) + min(abs(pg_t1[9] - pg_ctrl)) + min(abs(pg_t1[10] - pg_ctrl)))

sum(min(abs(pg_t2[1] - pg_ctrl)) + min(abs(pg_t2[2] - pg_ctrl)) + min(abs(pg_t2[3] - pg_ctrl)) + min(abs(pg_t2[4] - pg_ctrl)) + min(abs(pg_t2[5] - pg_ctrl)) + min(abs(pg_t2[6] - pg_ctrl)) + min(abs(pg_t2[7] - pg_ctrl)) + min(abs(pg_t2[8] - pg_ctrl)) + min(abs(pg_t2[9] - pg_ctrl)) + min(abs(pg_t2[10] - pg_ctrl)))



```

Using this metric, treatment 2 is closer to control. 


Finally, I'll consider the Jaccard similarity index of each treatment group versus control. To apply this to one-dimensional values, I will consider the range of each treatment as $| U \cap V |$, while the range of control and treatment will be $| U \cup V |$. Subtracting this value from 1 will give the Jaccard distance.


```{r, ex1-jaccard}

1- (max(pg_t1) - min(pg_t1)) / (max(cbind(pg_ctrl, pg_t1)) - min(cbind(pg_ctrl, pg_t1)))

1- (max(pg_t2) - min(pg_t2)) / (max(cbind(pg_ctrl, pg_t2)) - min(cbind(pg_ctrl, pg_t2)))

```

The Jaccard distance for treatment 1 is closer to control.

__Ex. 2__ -- Now use a higher-dimensional set _mtcars_, try the same distance metrics in the previous question and discuss the results. 


For a larger dimension dataset, I'll be using the built-in _dist_ function in R to calculate the distance. While _mtcars_ contains all numerical values for each variable, there is one dummy variable (transmission is 0 or 1), and several categorical variables (number of forward gears, carburetors, cylinders). I will consider the distance of each value 

```{r, mtcars-euclid}


sum(dist(as.matrix(mtcars), method = 'euclidian'))


```

Without summing the values, the _dist_ function would provide a matrix giving the distance of each car from each other car. I summed the resulting matrix for all of its values for easier comparison between distance metrics.

```{r, mtcars-manhattan}


sum(dist(as.matrix(mtcars), method = 'manhattan'))


```


For Jaccard distance, I used the _distance_ function as part of the 'ecodist' package.

```{r, mtcars-jaccard}


sum(ecodist :: distance(as.matrix(mtcars), method = 'jaccard'))


```

Compared to the 1-dimensional example, there appears to be much more variation between distance metrics. This comparison may underscore the importance of scaling initial values for effective use of distance metrics. 

__Ex. 3__ -- Use the built-in dataset _mtcars_ to carry out hierarchy clustering using two different distance metrics and compare if they get the same results. Discuss the results. 

```{r, mtcars-hclust}

plot(hclust(dist(scale(mtcars), method = 'euclidian')), main = 'Hierarchy Clustering, Euclidian Distance')
rect.hclust(hclust(dist(scale(mtcars), method = 'euclidian')), k = 3)

plot(hclust(dist(scale(mtcars), method = 'manhattan')), main = 'Hierarchy Clustering, Manhattan Distance')
rect.hclust(hclust(dist(scale(mtcars), method = 'manhattan')), k = 3)

```

After scaling data, the two distance metrics offered seemingly different clusters. I added calls to _rect_ to better visualize the differences. It appears that some of the clusters have similarities but there are discernible differences between the distance metrics.


__Ex. 4__ -- Load the well-known Fisher’s _iris_ flower dataset that consists of 150 samples for three species (50 samples per species). The four measures or features are the lengths and widths of sepals and petals. Use the kNN clustering to analyze this iris dataset by selecting 120 samples for training and 3 samples for testing. 


```{r, iris-knn}
set.seed(1114)

train_index <- sample(seq_len(150), size =120) #generate random index separating iris into specified samples
train <- iris[train_index, -5] #training data
test <- iris[-train_index, -5] #test data

train_cat <- iris[train_index, 5] 
test_cat <- iris[-train_index, 5]  #true designations for test

iris_predict <- (knn(train = train, test = test, cl = train_cat, k = 3)) #find predicted categories

table(iris_predict, test_cat) #contingency table

```

Using kNN only missed a single categorization at k = 3.



__Ex. 5__ -- Use the _iris_ dataset to carry out k-means clustering. Compare the results to the actual classes to estimate the clustering accuracy. 

 
```{r, iris-kmeans}
set.seed(1114)

iris_kmeans <-  kmeans(iris[,-5], centers = 3)$cluster #obtain cluster designations

round(table(iris$Species, iris_kmeans)*100/150,1) #create contingency table for cluster numbers versus species

```
 After dividing by n = 150 to obtain a percentage, it appears that clustering accuracy for k-means is not as high as knn. The _setosa_ species is present in clusters 2 and 3 at a 2:1 ratio, while _versicolor_ and _virginica_ are lumped together in cluster 1. Let's see if increasing number of centroids to 5 will improve accuracy

```{r, iris-kmeans5}
set.seed(1114)

iris_kmeans <-  kmeans(iris[,-5], centers = 5)$cluster #obtain cluster designations

round(table(iris$Species, iris_kmeans)*100/150,1) #create contingency table for cluster numbers versus species

```

While the result is a little difficult ot understand, it still appears there are accuracy problems with k-means compared to knn.
